\chapter{Publications During Graduate Study}

\noindent
1. S. Tristram-Nagle, R. Zhang, R.~M. Suter, C.~R. Worthington, 
{\bf W.-J. Sun}, and J.~F. Nagle, {\it Measurement of chain tilt angle in fully 
hydrated bilayers of gel phase lecithins}, Biophys. J. {\bf 64}, 1097-1109 
(1993).\\

\noindent
2. {\bf W.-J. Sun}, R. M. Suter, M. A. Knewtson, C. R. Worthington, 
S. Tristram-Nagle, R. Zhang, and J. F. Nagle, {\it Order and disorder in fully 
hydrated unoriented bilayers of gel phase dipalmitoylphosphatidylcholine},
Phys. Rev. E {\bf 49}, 4665-4676(1994).\\

\noindent
3. S. Tristram-Nagle, R. M. Suter, {\bf W.-J. Sun}, and J. F. Nagle,  {\it
Kinetics of subgel formation in DPPC: x--ray diffraction proves
nucleation-growth hypothesis}, Biochim. Biophys. Acta. {\bf 1191}, 14 (1994).\\

\noindent
4. R. Zhang, {\bf W.-J. Sun}, S. Tristram-Nagle, R. L. Headrick, 
R. M. Suter and J. F. Nagle, {\it Critical fluctuations in membranes}, 
Phys. Rev. Lett. {\bf 74}, 2832-2835 (1995).\\

\noindent
5. R. Zhang, S. Tristram-Nagle, {\bf W.-J. Sun}, R. Headrick, T. C. Irving, 
R. M. Suter, and J. F. Nagle, {\it Small Angle X-ray Scattering From Lipid 
Bilayers Is Well Described By Modified Caill\'e Theory, But Not By 
Paracrystalline Theory}, Biophys. J. {\bf 70} 349-357 (1996).\\

\noindent
6. J. F. Nagle, R. Zhang, S. Tristram-Nagle, {\bf W.-J. Sun}, H. Petrache, and
R. M. Suter, {\it X-ray structure determination of fully hydrated L$_{\alpha}$
phase DPPC bilayers}, Biophys. J. {\bf 70} 1419-1431 (1996).\\

\noindent
7. {\bf W.-J. Sun}, S. Tristram-Nagle, R. M. Suter, and J. F. Nagle, {\it 
Anomalous phase behavior of long chain saturated lecithin bilayers}, 
Biochim. Biophys. Acta. {\bf 1279}, 17 (1996).\\

\noindent
8. {\bf W.-J. Sun}, S. Tristram-Nagle, R. M. Suter, and J. F. Nagle, 
{\it Structure of the ripple phase in lecithin bilayers}, submitted to
the Proceedings of the National Academy of Sciences.\\

\noindent
9. {\bf W.-J. Sun}, S. Tristram-Nagle, R. M. Suter, and J. F. Nagle, {\it 
Structure of gel phase saturated lecithin bilayers: temperature and chain 
length dependence}, submitted to the Biophysical Journal.
