% ******************* Acknowledgements *********************
\prefacesection{Acknowledgements}

First and the most, I would like to thank my advisor, 
Prof. John F. Nagle, for his indispensable guidance, support, 
patience and tolerance. What I have learned from him during my Ph.D. 
time are multitudinous and will be beneficial to my future career 
and life.

I would like to thank Prof. Robert M. Suter for his unfailing
expertise and support on x-ray scattering experiments. When things
went awry, he was always there and ready to set things straight. 
His insightful comments and encouragements are highly appreciated.

I would also like to thank Dr. Stephanie Tristram-Nagle. It was she
who helped me make the very first step in the lab. Her devotion
and rigorous attitude towards science are always inspiring.
Beside the research duty, she has taken on by herself the responsibility 
of teaching and correcting my English. If I can speak and write better
English today, she deserves most of the credit.

I thank my fellow graduate students, Dr. Ruitian Zhang (now a postdoctoral
fellow at Ohio State University) and Mr. Horia Petrache. 
We shared most of the delightful and frustrating moments in
scientific research and in pursuit of high academic degrees.
Their friendship and help are appreciated. 

I also thank Dr. Kechuan Tu at University of Pennsylvania. The collaboration
and communication between us are highly delightful and fruitful.

I want to thank my wife, Ms. Meijing Peng, for her selfless support, 
encouragement and understanding.
 
Last but not the least, I would like to thank my parents for bringing me
to this world and their endless love, without which nothing is possible.
